Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-13 documents/ms.tex
minor wording

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process (GP) with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

We apply the ACF method to the sample of \naigrain\ simulated light
curves.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line,
more fall slightly below than above it, \ie\
rotation periods are preferentially underestimated.
This stems from how peak positions in the ACF are measured.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left
(towards shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.
We adopt this standard practice here to faithfully compare
our new method to that used in the literature.

@@

In order to test our rotation period inference method on real data,
we apply it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux
(\texttt{pdcsap\_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted, and mask out all known transiting planet candidate signals.
As with the simulated light curves, we randomly subsample each
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data, using the ACF-based prior as before.

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, with only a few
exceptions, demonstrating that this method works not only on simulated data,
but also on real data---with the caveat that for any particular data set, some
care is needed regarding the setting the priors and ranges for the GP
hyperparameters.
Notably, below $\ln P$ of about 2.5, the GP method recovers periods
systematically slightly larger than \citet{Mcquillan2013}---a likely consequence
of correcting for the ACF peak measurement bias discussed in Section \ref{sec:acf}.

>>>>

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process (GP) with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkoimcq\ \Kepler\ stars with 
previously measured periods.
Because this method delivers posteriors PDFs, it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

We apply the ACF method to the sample of \naigrain\ simulated light
curves.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line,
more fall slightly below than above it, \ie\
rotation periods are systematically underestimated.
This stems from how peak positions in the ACF are measured.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left
(towards shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.
We adopt this standard practice here to faithfully compare
our new method to that used in the literature.

@@

In order to test our rotation period inference method on real data,
we apply it to a set of \nkoimcq\ \Kepler\ Object of Interest (KOI)
host stars, that had rotation periods previously measured by
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux
(\texttt{pdcsap\_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted, and mask out all known transiting planet candidate signals.
As with the simulated light curves, we randomly subsample each
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data, using the ACF-based prior as before.

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, with only a few
exceptions, demonstrating that this method works not only on simulated data,
but also on real data---with the caveat that for any particular data set, some
care is needed regarding the setting the priors and ranges for the GP
hyperparameters.
Notably, the GP method recovers periods
systematically slightly larger than \citet{Mcquillan2013}---a likely consequence
of correcting for the ACF peak measurement bias discussed in Section \ref{sec:acf}.
